Two Decades of Smalltalk VM Development

### 2018-08-16 paper.tex
changes on introduction

\egb{I rewrote and shortened the first paragraph since the abstract was quite unbalance towards the research context but very scarce on details about the research problem and solution. I left the original text in comments below}

\openSmalltalkVM is a VM for languages in the Smalltalk family which is itself written in a subset of Smalltalk that can easily be translated to C.  Development is done in Smalltalk, an activity we call ``Simulation''. The production VM is derived by translating the core VM code to C. As a result, two execution models coexist: simulation, where the Smalltalk code is executed on top of a Smalltalk VM, and production, where the same code is compiled to executable code through the C compiler. While simulation is used to develop and debug the VM, production is used to release the VM.

@@

%TODO egb: revisit this paragraph after reading all paper.
In this paper, we detail the VM simulation infrastructure and we report our experience developing and debugging the VM within it. We mention some of the limitations and how we worked around them. We discuss specifically how we use the VM simulator to develop and debug two core VM components, the garbage collector and the just-in-time compiler. Then, we discuss how we use the simulation infrastructure to perform analysis on the runtime, directing some design decisions we have made to tune VM performance.

@@

To specify the the virtual machine (VM), the Smalltalk-80 team at Xerox PARC wrote a Smalltalk VM entirely in Smalltalk \cite{blueBook}. In 1995, members of the same team built an open-source Smalltalk and VM, Squeak \cite{SqueakByExample} using the code from \cite{blueBook} as a starting point. Part of the code base was however narrowed down to a subset of Smalltalk, called \emph{Slang}, to allow Smalltalk to C compilation \cite{BackToTheFuture}.  Dummy Smalltalk message sends were used to embed meta information in the code, such as \emph{self var: 'foo' type: 'char *'} which have no effect during simulation but guide the translation process. Effectively, the VM could be both simulated by executing the Slang code on top of the Smalltalk VM or compiled through the C compiler to native code to generate the production VM. At this point, the VM consisted mainly in:
\begin{itemize}
\item an interpreter with a spaghetti stack, 
\item a memory manager with a compact but complex object representation, a pointer-reversing tracing garbage collector and a heap divided into two generations
\item WarpBlt, a rotation and scaling extension for the bit-based BitBLT graphics engine.
\item external C code to support running the VM on popular platforms
\end{itemize} 
The first three parts of the VM were written entirely in Slang. A few extra features, such as file management, were written both in Smalltalk for simulation purposes and in C for the production VM.

Over the years, the Squeak VM evolved to give birth recently to \openSmalltalkVM\footnote{https://github.com/OpenSmalltalk/opensmalltalk-vm/}, the default VM for various Smalltalk and Smalltalk-like systems such as Pharo \cite{PharoByExample}, Squeak \cite{SqueakByExample}, Cuis, Croquet and Newspeak \cite{NewspeakOopsla}. As the VM evolved, the simulator co-evolved as a tool to develop and debug the VM. One change moved the meta data that guides translation out of the executable Smalltalk into pragmas \cite{Pragmas}.  The most significant evolution of the simulator came with the introduction of the Just-In-Time compiler (JIT). The code base was liberally sprinkled with assertions.  The spaghetti stack was mapped to a more conventional stack frame organisation in a stack zone of a few hundred k bytes organised into small pages, a scheme called context-to-stack mapping \cite{EfficientST80}, \cite{StackMappingVW}.  Machine code generated by the JIT was executed by binding multiple processor simulators (first Bochs \cite{Bochs} for x86, SkyEye for ARMv6, Smalltalk code for 32-bit MIPS).  A new object representation and garbage collector was added to also support 64-bits \cite{Spur}, which required refactoring the interpreter to allow the object representation to be chosen at startup. Bochs was used a second time for x64 support.

>>>>

\egb{I rewrote and shortened the first paragraph since the abstract was quite unbalance towards the research context but very scarce on details about the research problem and solution. I left the original text in comments below, and I moved some details to the introduction were they were more necessary in my opinion.}

\openSmalltalkVM is a virtual machine (VM) for languages in the Smalltalk family (e.g. Squeak, Pharo) which is itself written in a subset of Smalltalk that can easily be translated to C.  Development is done in Smalltalk, an activity we call ``Simulation''. The production VM is derived by translating the core VM code to C. As a result, two execution models coexist: simulation, where the Smalltalk code is executed on top of a Smalltalk VM, and production, where the same code is compiled to executable code through the C compiler.

@@

%TODO egb: revisit this paragraph after reading all paper to refine problem statement and solution.
In this paper, we detail the VM simulation infrastructure and we report our experience developing and debugging the VM within it. We mention some of the limitations and how we worked around them. We discuss specifically how we use the VM simulator to develop and debug two core VM components, the garbage collector and the just-in-time compiler. Then, we discuss how we use the simulation infrastructure to perform analysis on the runtime, directing some design decisions we have made to tune VM performance.

@@

To specify the virtual machine (VM), the Smalltalk-80 team at Xerox PARC wrote a Smalltalk VM entirely in Smalltalk \cite{blueBook}. In 1995, members of the same team built an open-source Smalltalk \egb{dialect?} and VM, Squeak \cite{SqueakByExample}, using the code from \cite{blueBook} as a starting point. Part of the code base was, however, narrowed down to a subset of Smalltalk, called \emph{Slang}, to allow Smalltalk to C compilation \cite{BackToTheFuture}.

\egb{Rewrote the paragraph to add old info from the abstract and add some flow to the explanation. Maybe we should also define here what simulation and production is (since the abstract is not considered to be part of the main body of the paper)}

Effectively, the VM could be both simulated by executing the Slang code on top of the Smalltalk VM or compiled through the C compiler to native code to generate the production VM. 
The production VM can be derived by translating the core VM code to C, combining the generated C code with a set of platform-specific support files, and compiling with the platform's C compiler.
Dummy Smalltalk message sends were used to embed meta information in the code, such as \emph{self var: 'foo' type: 'char *'} which have no effect during simulation but guide the translation process of the executable Smalltalk.

At this point, the VM consisted mainly in:
\begin{itemize}
\item an interpreter with a spaghetti stack, 
\item a memory manager with a compact but complex object representation: a pointer-reversing tracing garbage collector and a heap divided into two generations,
\item WarpBlt, a rotation and scaling extension for the bit-based BitBLT graphics engine,
\item and external C code to support running the VM on popular platforms.
\end{itemize}

The first three components of the VM were written entirely in Slang. A few extra features, such as file management, were written both in Smalltalk for simulation purposes and in C for the production VM.

Over the years, the Squeak VM evolved to give birth recently to \openSmalltalkVM\footnote{https://github.com/OpenSmalltalk/opensmalltalk-vm/}, the default VM for various Smalltalk-like systems such as Pharo \cite{PharoByExample}, Squeak \cite{SqueakByExample}, Cuis, Croquet and Newspeak \cite{NewspeakOopsla}.

As the VM evolved, the original simulator co-evolved as a tool to develop and debug the VM. The meta data used to guide the translation process was replaced by \emph{pragmas} \cite{Pragmas}.  
%One change moved the meta data that guides translation out of the executable Smalltalk into pragmas \cite{Pragmas}.  
However, the most significant evolution of the simulator came with the introduction of the Just-In-Time compiler (JIT). 
%The code base was liberally sprinkled with assertions. 
The spaghetti stack was mapped to a more conventional stack frame organisation in a stack zone of a few hundred k bytes organised into small pages, a scheme called context-to-stack mapping \cite{EfficientST80}, \cite{StackMappingVW}.  Machine code generated by the JIT was executed by binding multiple processor simulators (first Bochs \cite{Bochs} for x86, SkyEye for ARMv6, Smalltalk code for 32-bit MIPS).  
Finally, a new object representation and garbage collector was added to also support 64-bits \cite{Spur}, which required refactoring the interpreter to allow the object representation to be chosen at startup. Bochs was used a second time for x64 support.

\egb{at this point as reading I have all the history and context to understand the major milestones in the development of Smaltalk VMs but I still do not know what this paper is about :P. You need to introduce now the goal of the paper or the problem you want to explain the reader you guys tackled. From the brief discussions I had last week with Clement, I was thinking that the paper will focus on explaining the tools that you guys built to help debugging and developing the VM. You don't have this focus currently in the text. I will continue reviewing to see if I can get it out the rest of the paper.}
